Analysing Constraint Grammars with a SAT-solver

### 2016-02-04 paper/lrec/implementation.tex
more stuff to paper

>>>>

\paragraph{Creating realistic ambiguities}

\todo{List of word forms+readings; list of distinct readings; make into matrix where we see that x and y can be ambiguous but a and z never; smart things with constraints (Koen)}

Now with this, we can detect even these kinds of rules as problematic:

 \begin{itemize}
 \item[] \begin{verbatim}
 REMOVE verb IF (-1 det) ;
 REMOVE noun IF (-1 det) ;
 \end{verbatim}
 \end{itemize}

 With our symbolic sentence, these rules will be no problem; to apply the latter, we only need to construct a target that has a realistic ambiguity with a noun; the verb is gone already.
 However, in the context of real life texts, the grammar writer might actually mean the following:

 \begin{itemize}
 \item[] \begin{verbatim}
 REMOVE verb IF (-1 det) (0 noun) ;
 REMOVE noun IF (-1 det) (0 verb) ;
 \end{verbatim}
 \end{itemize}

 In that case, those rules are contradictory, and we would like to find it out. The additional constraints will prevent the SAT solver from creating an ambiguity where noun is ambiguous with e.g. punctuation.

### 2016-02-04 paper/lrec/problem_desc.tex
more stuff to paper

Most of NLP research has been since the 90s aimed towards data-driven methods. While these approaches have produced good results for languages with large annotated corpora, they are practically unusable for low-resourced languages. Even for languages with many resources, \todo{cite Manning (2011)} argues that rule-based methods are needed to push the accuracy of POS taggers from the 97\%s to the near 100\%. POS tagging is a fundamental phase which is required in further syntactic or semantic analysis---hence a good POS-tagger is an important resource for any language. We believe that Constraint Grammar is a good candidate especially for languages with low resources, and in this paper we describe methods for CG writers and users to analyse and improve their grammars.

@@

CGs are valuable resources for rule-based NLP, especially for lesser
resourced languages. They are robust and can be written without large
corpora---only morphological analysis is needed. The formalism is
lightweight and language-independent, and resources can be shared
between related languages \cite{bick2006spanish,lene_trond_linda2010}.
%Mature CGs contain some thousands of rules, but even small CGs are shown to be effective \cite{lene_trond2011}.

As CGs grow larger, it gets harder for the grammar writers and users
to keep track of all the rules and their interaction.
With our tools, the users can find conflicting rules, diagnose problems and improve their grammars.
Some examples of such conflicts are listed below:
\begin{itemize}
\item If a rule appears twice, the second occurrence will be disabled by the first
\item $R$ selects something in a context, $R'$ removes~it
\item $R$ removes something from the context of $R'$, so $R'$ can never
  apply
\item $R$ has an internal conflict, such as non-existent
tag combination, or contradicting requirements for a context word
\end{itemize}
$R$ can also be a set of rules: for instance, if one rule removes a verb in
context $C$, and another in context $\neg C$, together these rules
remove a verb in all possible cases, disabling any future rule that
targets verbs.

>>>>

%Most of NLP research has been since the 90s aimed towards data-driven methods. While these approaches have produced good results for languages with large annotated corpora, they are practically unusable for low-resourced languages. Even for languages with many resources, \todo{cite Manning (2011)} argues that rule-based methods are needed to push the accuracy of POS taggers from the 97\%s to the near 100\%. POS tagging is a fundamental phase which is required in further syntactic or semantic analysis---hence a good POS-tagger is an important resource for any language. We believe that Constraint Grammar is a good candidate especially for languages with low resources, and in this paper we describe methods for CG writers and users to analyse and improve their grammars.

@@

CGs are valuable resources for rule-based NLP, especially for lesser
resourced languages. They are robust and can be written without large
corpora---only morphological analysis is needed. The formalism is
lightweight and language-independent, and resources can be shared
between related languages \cite{bick2006spanish,lene_trond_linda2010}.
Mature CGs contain some thousands of rules, but even small CGs are shown to be effective \cite{lene_trond2011}.

As CGs grow larger, it gets harder for the grammar writers and users
to keep track of all the rules and their interaction.
This is partly by design: CG is a shallow formalism, and individual rules can be written to address any property, 
%such as "any verb" or "the word form `foo' which is tagged as agent"

With our tools, the users can find conflicting rules, diagnose problems and improve their grammars. 
We expect two major use cases: 
one where the grammar writer takes a whole grammar and cleans it up, and
other where the grammar writer tests new rules as they write them.

%%%%% Motivate testing some rules, not whole grammar

Given the following $n$ rules, a grammar writer may ask some questions while writing a grammar. 
\begin{itemize}
\item[] \texttt{SELECT Inf IF (-1 Para OR De) (0C V) ;}
\item[] \texttt{SELECT Inf IF (-1 Pr) (0C V) ;}
\item[] \texttt{SELECT Inf IF (-1C Vai) ;}
\item[] \texttt{SELECT Inf IF (-1C Vbmod) (0C V) ;}
\item[] \texttt{SELECT Inf IF (-1C Ter/de) ;}
\item[] \texttt{SELECT Inf IF (-1C Vbmod) (0 Ser) ;}
\end{itemize}

\begin{itemize}
\item Are all the rules distinct? (e.g. Para and De are included in Pr)
\item Could two or more rules be merged (\texttt{IF -1 Pr OR Vai OR Vbmod ...})?
\item Does the order of the rules matter? What is the best order?
\end{itemize}

If there

%%%%%

Some examples of such conflicts are listed below:
\begin{itemize}
\item If a rule appears twice, the second occurrence will be disabled by the first
\item $R$ selects something in a context, $R'$ removes~it
\item $R$ removes something from the context of $R'$, so $R'$ can never
  apply
\item $R$ has an internal conflict, such as non-existent
tag combination, or contradicting requirements for a context word
\end{itemize}
$R$ can also be a set of rules: for instance, if one rule removes a verb in
context $C$, and another in context $\neg C$, together these rules
remove a verb in all possible cases, disabling any future rule that
targets verbs.
